---
title: 'MDI: Weeks 1 - 3'
author: Stephen Coleman
date: '2019-02-20'
slug: log-weeks-1-3
categories: []
tags:
  - Chris Wallace
  - MDI
bibliography: weeks_123_bib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prelude

Many of the R snippets below require these R packages [@kolde2019pheatmap, @dowle2019data, @neuwirth2014rcolorbrewer]:

```{r libraries}
library(pheatmap)
library(data.table)
library(RColorBrewer)
```

## Reading
Inital work involved reading @mason2016mdi and reading the manual of the associated program. Expecting that this would have difficulties scaling (based on Paul's beliefs from November) to nine datasets different sampling methods were researched [@bardenet2017markov]. Hamiltonian Monte Carolo appears a likely candidate to improve the rate of convergence, with @betancourt2017conceptual providing an explanation of the theoretical framework. These topics were of interest as the MDI implementation uses a Gibbs sampler. We expected that we could make gains in the computational tractability of using all 9 datasets in a single call of MDI with a more clever sampling method.


## Data handling
MDI uses data that has common row names across all datasets. The data from [the CEDAR cohort](http://139.165.108.18/srv/genmol/permanent/1be6993fe41c12a051c9244d67c91da2be49e5dd26a6cd79f442bc006971e2ef/crohn-index.html) is in the form of [people $\times$ probes]. We want to transpose this; to accomplish this task we wrote [an Rscript](https://github.com/stcolema/ra_chris_wallace/blob/master/matrix_transposer.R) that can be called form the Linux command line with appropriate arguments to transpose .csv files in a given directory. The `data.table` package by @dowle2019data was of great use in making this quick to run - for our 9 datasets of ~300 people and 8-16 thousand probes it took less than a minute to read, convert and write the files.

Another [script was required to remove all the non-ubiquitous probe IDs](https://github.com/stcolema/ra_chris_wallace/blob/master/common_rownames_data.R). Any probe not present in all datasets was removed to allow MDI to perform as MDI cannot handle missing data. This reduced the set of probes to 4,964.

We performed MDI on these 9 reduced datasets with 10,000 iterations, a burn-in of 1,000 and a thinning rate of 25. MDI overfits the clustering problem beginning with 50 clusters; this reduced down to around 10 occupied clusters per dataset. We assumed that the median cluster allocation was the predicted cluster forgetting about the **label-flipping** problem of unsupervised methods.

We constructed a dataframe of the median allocation per dataset for each probe. The probes are translated to the associated gene using the [key available online](http://139.165.108.18/srv/genmol/permanent/1be6993fe41c12a051c9244d67c91da2be49e5dd26a6cd79f442bc006971e2ef/CEDAR_GE/Probes_good_reanno_31137_TSS.txt). As some of the probes mapped to non-unique genes we use a matrix to hold our data (as this does not require unique row names). We read this in and visually inspect the head of the data.

```{r mdi_1_alloc_data}
# Read in the data and convert to a matrix with apporpriate row names
compare_df <- data.table::fread("allocation_data_run_1.csv")
compare_df_mat <- as.matrix(compare_df[, -10])
row.names(compare_df_mat) <- compare_df$V1

# Inspect the data
head(compare_df_mat)
```

## Visualisation

Using this data we generated a heatmap.

```{r mdi_1_heatmap, echo = FALSE}
col_pal <- c("#DDDDDD", rev(RColorBrewer::brewer.pal(n = 11, "RdYlBu")))

ph_full <- pheatmap::pheatmap(compare_df_mat,
 cluster_cols = T,
 color = col_pal,
 main = "Heatmap for common probes"
)

```

Notice that we have clusters of datasets; CD4, CD8 and CD14, RE and IL and then CD15, CD19 and TR. Platelets are outside of any clusters.

There are 14 clusters present here; to enable comparison we went for a wider colour palette. This is the reason for the garishness of this heatplot. There is appears to have lower similarity than one might expect, particularly considering the selection of data - we only include the probes that are sufficiently expressed in each cell type to register for measurement. Possibly this is an artifact of not considering label-flipping. 

### Investigate subsections

With the same allocation data and maintaining the ordering from the heatmap of the full dataset we inspect subsamples of the data.

```{r mdi_1_small_heatmaps, echo = FALSE}
row_order <- ph_full$tree_row[["order"]]

df_ph_order <- compare_df_mat[row_order, ]

pheatmap(df_ph_order[1:1000, ],
         cluster_rows = F, 
         cluster_cols = F,
         color = col_pal,
         main = "Entries 1:1000")

pheatmap(df_ph_order[1001:2000, ],
         cluster_rows = F, 
         cluster_cols = F, 
         color = col_pal,
         main = "Entries 1001:2000")

pheatmap(df_ph_order[2001:3000, ], 
         cluster_rows = F, 
         cluster_cols = F, 
         color = col_pal,
         main = "Entries 2001:3000")

pheatmap(df_ph_order[3001:4000, ], 
         cluster_rows = F, 
         cluster_cols = F, 
         color = col_pal,
         main = "Entries 3001:4000")

pheatmap(df_ph_order[4001:4964, ], 
         cluster_rows = F,
         cluster_cols = F, 
         color = col_pal,
         main = "Entries 4001:4964")

```

With this increased granularity, we see that after the first 1,000 rows the similarity across tissues becomes greater. There is still large sections of dissimilarity, but the clustering more coherent than the first plot suggests. Notice also that the PLA dataset (platelets) is the greatest source of dissimialrity - this is also the thinnest dataset with only 8 thousand probes present in the full format. 

### Investigate different combinations

Please note that these following heatmaps have **unique** clusterings and have not inherited the clustering of the original heatmap unlike those preceding this section.

```{r no_platelets, echo = FALSE}
pheatmap::pheatmap(compare_df_mat[, -7],
 cluster_cols = F,
 color = col_pal,
 main = "Heatmap for common probes excluding PLA"
)
```

We consider only the most similar datasets based on the clustering of datasets: CD4, CD8, CD14, RE and IL.
 
```{r sim_heatmaps, echo = FALSE}
pheatmap::pheatmap(compare_df_mat[, -c(2, 6, 7, 9)],
 cluster_cols = T,
 color = col_pal,
 main = "Heatmap for datasets of greatest similarity"
)
```

Out of curiousity we check the clustering and heatmap for the excluded datasets.

```{r dis_heatmaps, echo = FALSE}
pheatmap::pheatmap(compare_df_mat[, c(2, 6, 7, 9)],
 cluster_cols = T,
 color = col_pal,
 main = "Heatmap for datasets of least similarity"
)
```

These are more coherent than our first heatmap where we considered all datasets at once - this suggests that trying to process the output of all 9 datasets at once is not optimal and is in itself a non-trivial task (particulalrly if label-flipping cannot be ignored).

## Including all probes

Next we attempt to include all probes present across the 9 datasets. The challenge here is to include the probes in such a fashion as to not add noise to the final clustering. To do this we first inspect the spread of the expression values in the datasets. As this is too large to consider in normal summary statistics, we inspect histograms of these - we notice that all of the probe expression values have very similar means and standard deviations - these are densely (i.e. within 1.0 for the mean, within 0.2 for the standard deviation) distributed around 7.0 and 2.10 respectively.

We created [an Rscript](https://github.com/stcolema/ra_chris_wallace/blob/master/finad_all_probe_ids.R) that finds the list of all unique probe IDs present across the 9 datasets and adds the required rows to each .csv file (filled with 0's but any arbitrarily distant number from the range inhabitated by the actual measurements should be acceptable) and rearranging to have common row names and row orders in each dataset for all 18,517 prob IDs. Using these new datasets we perform MDI with the same arguments.

This script also saves a binary matrix indicating for each probe which datasets it is present in. We use this as part of a sense check to ensure that the cluster containing the imputed probes contains no other members. This is done by taking the Hadamard product of the binary matrix with the allocation matrix and comparing the clusters occupied in this to those of the allocation matrix. Ideally one of the clusters should have completed disappeared and been replaced by a 0 cluster. This is the case below.

As in the initial attempt we feed in our data and move the to a matrix format.

```{r mdi_2_alloc_data}
# Read in the data and convert to a matrix with apporpriate row names
compare_df_all <- data.table::fread("allocation_data_run_2.csv")
compare_df_all_mat <- as.matrix(compare_df_all[, -10])
row.names(compare_df_all_mat) <- compare_df_all$V1
 
```

We now create a heatmap of these genes.

```{r mdi_2_heatmap, echo = FALSE}
col_pal <- c("#DDDDDD", rev(RColorBrewer::brewer.pal(n = 11, "RdYlBu")))
ph_full <- pheatmap::pheatmap(compare_df_all_mat,
  cluster_cols = T,
  color = col_pal,
  main = "Heatmap for all genes"
)

row_order <- ph_full$tree_row[["order"]]

df_ph_order <- compare_df_all_mat[row_order, ]

```

Notice the clustering of datasets; we have groupings of the intestinal sample cells (IL, RE and TR), of CD14, CD19, CD4 and CD8 and then CD15 and platelets on their own.

The grey areas are absent genes for the dataset in question. This is too dense (being some 18,517 observations long), so we consider subsets. 

```{r mox_pheatmap, echo = FALSE}
MOX_ind <- grep("MOX", row.names(compare_df_all_mat))
pheatmap(compare_df_all_mat[MOX_ind, ],
         cluster_rows = F,
         cluster_cols = F,
         color = col_pal
)

```

Unfortunately too many of the MOX genes are absent from too many datasets to read
to significantly into these results.

Investigate higher granularity.

```{r mdi_2_granularity, echo = FALSE}
pheatmap(df_ph_order[1:2500, ],
  cluster_rows = F,
  cluster_cols = F,
  color = col_pal,
  main = "Heatmap for genes 1:2500"
)

pheatmap(df_ph_order[2501:5000, ],
  cluster_rows = F,
  cluster_cols = F,
  color = col_pal,
  main = "Heatmap for genes 2501:5000"
)

# There's very little information here
pheatmap(df_ph_order[5001:10000, ],
  cluster_rows = F,
  cluster_cols = F,
  color = col_pal,
  main = "Heatmap for genes 5001:10000"
)

pheatmap(df_ph_order[10001:15000, ],
  cluster_rows = F,
  cluster_cols = F,
  color = col_pal,
  main = "Heatmap for genes 10001:15000"
)

# Here it quite similar
pheatmap(df_ph_order[15001:18517, ],
  cluster_rows = F,
  cluster_cols = F,
  color = col_pal,
  main = "Heatmap for genes 15001:18517"
)
```

Recalling the clustering of datasets from the initial heatmap, we investigate these groupings. We inspect different combinations trying the CD combinations and the intestinal combinations.

```{r pheatmap_combinations_mdi_2, echo = FALSE}
pheatmap::pheatmap(compare_df_all_mat[, c(1, 3, 4, 5)],
  cluster_cols = T,
  color = col_pal,
  main = "Heatmap for all genes"
)

pheatmap::pheatmap(compare_df_all_mat[, c(6, 8, 9)],
  cluster_cols = T,
  color = col_pal,
  main = "Heatmap for all genes"
)

```

To show that platelets and CD15 do not sit natually in any clustering of datasets we include the following heatmaps. The sparseness of the platelets dataset causes confusion and CD15 does not sit naturally in either clustering.

```{r pheatmap_combinations_mdi_2_pla_cd15, echo = FALSE}
pheatmap::pheatmap(compare_df_all_mat[, c(7, 2, 1, 3, 4, 5)],
  cluster_cols = T,
  color = col_pal,
  main = "Heatmap for all genes"
)

pheatmap::pheatmap(compare_df_all_mat[, c(2, 7, 6, 8, 9)],
  cluster_cols = T,
  color = col_pal,
  main = "Heatmap for all genes"
)
```

# References
